Neural Connectome of the Ctenophore Statocyst

### 2025-04-04 manuscript/Manuscript.qmd
MS: edits
MS and correlation plot: edits

---
title: "Neural Circuit Organization of the Ctenophore Aboral Organ"

subtitle: |
  **Kei Jokura**$^{1,2,3,4*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewalow**$^{6}$, **Pawel Burkhardt**$^{5}$, **Gáspár Jékely**$^{1,2,6,*}$  
  <br><br>  
  $^{1}$ Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom  
  <br>
  $^{2}$ Biosciences, Faculty of Health and Life Sciences, University of Exeter, Exeter EX4 4QD, UK
  <br>
  $^{3}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
  <br>  
  $^{4}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
  <br>  
  $^{5}$ Michael Sars Centre, University of Bergen, Norway  
  <br>
  $^{6}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  <br><br>  
  **$^{*}$ Correspondence:**  
  [jokura@nibb.ac.jp](mailto:jokura@nibb.ac.jp),  
  [gaspar.jekely@cos.uni-heidelberg.de](mailto:gaspar.jekely@cos.uni-heidelberg.de)
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
max-width: 100%
format:
  html:
    embed-resources: true 
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: false
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

@@

Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth. During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns. To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*. We reconstructed 909 cells, including syncytial neurons that form a nerve net and synapse on the balancer cells. Our connectome analysis revealed that these syncytial neurons establish direct and indirect pathways to balancer cells, with "bridge cells" acting as relay neurons. High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation. Given that ctenophores branched off early from other animals, these results inform our understanding of the early divergence of nervous systems.

@@

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synalse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei This neuron has X <!--# add number --> presynaptic sites. The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and with two nuclei, and each containing X presynaptic structures. <!--#  -->

@@

@@

@@

We conducted the analysis using the cydippid stage of *M. leidyi* at five days post-fertilization. A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass. Filtered natural seawater was introduced under the coverslip, along with the cydippids. By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids. To stabilize the statolith's position, the microscope was tilted 90 degrees, arranging the stage vertically. The movement of the balancer cilia was observed using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS). Observations were made using a 40x objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion. For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals. The camera and LED strobe system were operated using the Video Capture Software BohNavi. Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED. To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.

>>>>

---
title: "Neural Connectome of the Ctenophore Statocyst"
subtitle: |
  **Kei Jokura**$^{1,2,3,4*}$, **Sanja Jasek**$^{1,2,6}$, **Lara Kewalow**$^{6}$, **Pawel Burkhardt**$^{5}$, **Gáspár Jékely**$^{1,2,6,*}$  
  <br><br>  
  $^{1}$ Living Systems Institute, University of Exeter, Exeter, EX4 4QD, United Kingdom  
  <br>
  $^{2}$ Biosciences, Faculty of Health and Life Sciences, University of Exeter, Exeter EX4 4QD, UK
  <br>
  $^{3}$ Exploratory Research Center on Life and Living Systems (ExCELLS), Okazaki, 444-8787, Japan  
  <br>  
  $^{4}$ National Institute for Basic Biology (NIBB), Okazaki, 444-8585, Japan  
  <br>  
  $^{5}$ Michael Sars Centre, University of Bergen, Norway  
  <br>
  $^{6}$ Heidelberg University, Centre for Organismal Studies (COS), 69120 Heidelberg, Germany  <br><br>  
  **$^{*}$ Correspondence:**  
  [jokura@nibb.ac.jp](mailto:jokura@nibb.ac.jp),  
  [gaspar.jekely@cos.uni-heidelberg.de](mailto:gaspar.jekely@cos.uni-heidelberg.de)
mainfont: Arial
monofont: DejaVu Sans Mono #Times New Roman
fontsize: 11pt
linestretch: 1.2
linkcolor: blue
documentclass: article
max-width: 100%
format:
  html:
    embed-resources: true 
  pdf:
    pdf-engine: xelatex
    number-sections: false
    margin-left: 0.8in
    margin-right: 0.8in
    margin-top: 0.8in
    margin-bottom: 0.8in
    toc: false
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{lineno}
    include-before-body:
      text: |
        \linenumbers
bibliography: references.bib #add bibliography in this file
csl: elife.csl #change according to journal
---

@@

Ctenophores possess a unique gravity receptor (statocyst) in their aboral organ formed by four clusters of ciliated balancer cells that collectively support a statolyth. During reorientation, differential load on the balancer cilia leads to altered beating of the ciliated comb rows to elicit turns. To study the neural bases of gravity sensing, we imaged by volume electron microscopy (vEM) the aboral organ of the ctenophore *Mnemiopsis leidyi*. We reconstructed 909 cells, including syncytial neurons that form a nerve net. The syncytial neurons synapse on the balancer cells and also form reciprocal connection with the bridge cells that span the statocyst. High-speed imaging revealed that balancer cilia beat and arrest in a coordinated manner but with differences between the sagittal and tentacular planes of the animal, reflecting nerve-net organisation. Our results suggest a coordinating rather than effector function for the nerve net and inform our understanding of the diversity of nervous-systems organisation across animals.

@@

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei This neuron has X <!--# add number --> presynaptic sites. The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and with two nuclei, and each containing X presynaptic structures. <!--#  -->

@@

<!--# discuss frequency data -->

@@

<!--#  discuss that it is a complete circuit, with no neuronal connections in or out - effector output is likely by hydrodynamics, as suggested by Tamm -->

@@

We conducted the analysis using the cydippid stage of *M. leidyi* at five days post-fertilization. A coverslip with a thin layer of Vaseline applied to two edges was lightly placed on a slide glass. Filtered natural seawater was introduced under the coverslip, along with the cy3dippids. By gently moving the coverslip, the desired orientation for observation was adjusted, and once positioned, slight pressure was applied to immobilize the cydippids. To stabilize the statolith's position, the microscope was tilted 90 degrees, arranging the stage vertically. The movement of the balancer cilia was observed using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a highly sensitive CMOS monochrome camera optimized for the near-infrared (NIR) range (UI-3360CP-NIR-GL Rev.2, iDS). Observations were made using a 40x objective lens (Objective LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27) with glycerine immersion. For recording, we employed a custom-made NIR LED strobe illumination system (wavelength: 850 nm) synchronized with the camera's exposure signals. The camera and LED strobe system were operated using the Video Capture Software BohNavi. Images were recorded at a resolution of 640×480 pixels with a frame rate of 100 fps for 2 minutes, using 0.05 ms pulses from the LED. To analyze the ciliary beating of the balancer cilia, we utilized the Multi Kymograph project in Fiji software.
